Elevational range shifts of mountain bumble bees and their floral hosts will be punctuated by nonlinear turnover of species and interactions

### 2021-02-27 ms_1/ms_1.Rmd
back up

---
title: "Bumble bees and wildflowers along a mountain slope exhibit nonlinear patterns of abundance and turnover, punctuated by the tree line ecotone"

@@

@@

The rate of modern climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Species range shifts, however, are not merely abstract, Cartesian translations along axes of latitude, elevation, and temperature; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Moreover, some of these factors, unlike temperature, cannot be assumed to vary linearly with latitude or elevation but instead exhibit marked nonlinearities and thresholds. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir]. Thus, species range shifts --- and, perhaps more importantly, the ecological consequences thereof --- are contingent upon patterns of species distributions and biotic interactions, and community ecology is as much prior as posterior to the ecology of global change.

Bumble bees (Hymenoptera: *Bombus* spp.) are both a classic model organism in community ecology and an emergent conservation priority due to recent declines [@Goulson2008-oe] and manifest vulnerability to climate change [@Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya), and its subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, bumble bees remain characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity associated with mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective, but they also provide a model system in which to study the general phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, necessitate an understanding the current elevational and temporal patterns of bumble bee communities and those of their floral hosts that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we move to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most acute.

@@

The study was conducted in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management. The sites sampled in this study are the same as those used in a series of previous studies [@Hoiss2012-sn; @Hoiss2012-kr; @Hoiss2015-gv], but the data reported here are otherwise independent.

Sampling consisted of repeated visits to each study site at approximately weekly intervals. Samples were only collected on dry days when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on or a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers.

In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.1 m^2^. Species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date (excluding males). To represent floral abundance from the perspective of each bumble bee species in our network, we first scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of visited genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp].

@@

All analyses were conducted in `R` `r citep(citation())`. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa], and the models we fit correspond to the “type I” model form described in @Pedersen2019-kf. PCoA and species β-diversity calculation were performed with the package `vegan` `r citep(citation("vegan"))`. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` `r citep(citation("bipartite"))`. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` `r citep(citation("gdm"))`. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("gdm"))`. Annotated R code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species groups lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

@@

The remaining species exhibited relatively diffuse patterns of abundance. Of these, *B. soroensis* was exceptional in that it exhibited peak abundance at the highest sites, but it was also found at relatively high abundance across the whole elevation gradient, peaking at mid-season. *B. hortorum* and *B. lapidarius* peaked at low elevation. *B. jonellus* exhibited little elevational variation in abundance until declining above 1500 m, but its abundance was strongly patterned in time, peaking early in the season. *B. terrestris/lucorum* was generally abundant across the whole elevation gradient and throughout the whole season, but it showed a weakly trimodal pattern in elevation, with abundance peaks at low, middle, and high sites. Its strongest abundance peak occurred at low elevation, early in the season during colony-founding. *B. wurflenii* was broadly abundant above 1000 m but was rarer at lower sites.

Elevation-time GAM smooths of floral abundance differed significantly from a flat surface for all species (p << 0.00001) and explained 64.8% of total deviance (adjusted R^2^ = 0.425). Floral resource availability exhibited similar patterns for most bumble bee species, with peak floral resources occurring below 1500 m and during mid-season **(Figure 2)**. There was a noticeable temporal lag in floral resource availability associated with increasing elevation, consistent with expected temperature constraints on plant phenology. The only species to exhibit marked anomalies in floral resource availability was *B. gerstaeckeri*, which specializes on the genus *Aconitum*[@Ponchau2006-af] but was also observed (presumably collecting nectar) on *Gentiana*, *Salvia*, *Rhinanthus*, *Cirsium*, and *Clinopodium.* Its floral resources were concentrated in two discrete peaks, the first at low elevation and mid-season, and the second at high elevation and late-season.

@@

```{r fig4, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to geographic, elevational, and temporal distance between site-dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient (i.e. geographic distance, elevation, or day of year), with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

@@

- Recap the point of the study

@@

The most salient pattern in our study is the unique importance of the elevation band falling at about 1500 m, which corresponds to the lower edge of the tree line ecotone in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn] and specifically that of the elevational transects used in our study. The floristic importance of the tree line ecotone has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id]. By studying the floral community through the lens of foraging bumble bees, however, we provide a functional extension of this floristic pattern, demonstrating that the tree line ecotone comprises (1) a boundary above which floral resource availability sharply declines, (2) a distributional interface between high-elevation bumble bees and their low-elevation or habitat-generalist counterparts, and (3) a zone of rapid acceleration of species and interaction β-diversity. The significance of the tree line ecotone as a threshold of floral resource availability and as a zone of overlapping bumble bee distributions corroborates the recent findings of @Minachilis2020-xd in the Mediterranean system of Mount Olympus, indicating that these patterns are likely to generalize well across mountain ecosystems. Our finding that interaction β-diversity is also punctuated by the tree line ecotone follows naturally from these compositional patterns, and we expect that the β-diversity patterns found in our system will prove similarly generalizable.

The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests a danger that high elevation bumble bees could move upslope faster than their floral resources. Indeed, a somewhat unexpected finding of our study is that high-elevation bumble bee species (*B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus*) were associated with floral hosts that peaked in abundance at much lower elevations, largely outside the range of their bumble bee visitors. As the upper boundary of high floral resource abundance, the lower edge of the tree line ecotone likely functions as critical foraging habitat for high-elevation bumble bees. Future warming could force these species farther into resource-poor alpine grasslands while at the same time increasing resource competition at the tree line as dominant low-elevation species, like *B. pascuorum*, extend their ranges upslope.

Our analyses of species and interaction β-diversity reveal a striking contrast between a relative stable bumble bee community and an extremely dynamic floral community. The dominance of total interaction β-diversity by floral species turnover is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps a predictable relationship given the exceptional cold-hardiness of bumble bees [@Heinrich1994-cv] and the sensitivity of plants to temperature gradients []. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history [] and that may explain the remarkable flexibility and intelligence of their flower handling []: how to forage efficiently in a diverse that turns over rapidly in both space and time. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df.

@@

- What do these findings tell use about climate change and range shifts in mountain ecosystems more generally?
- Do our results suggest any management strategies?

As in the past, so now are mountains the primarily theaters of ecological change for bumble bees.

Mountains unique ecological theaters in which extrinsic temporal climate change interacts with an intrinsic elevational climate gradation to alter the spatial and temporal distributions of species and interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj].

It should also be expected that climate change on mountain slopes will be punctuated by the tree line ecotone, which constitutes, at least with respect to bumble bees and wildflowers, a region of unique resource abundance and a transition to high and accelerating turnover of species and interactions [@@Descombes2017-id].

though as a matter of applied conservation it is perhaps doubtful that local interventions could forestall for long the consequences of a process occurring at a global scale.

Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees

The fact that

As such, the tree line ecotone deserves special consideration in conservation management, and targeted studies of plant-pollinator relationships at the tree line ecotone would be warranted.

>>>>

---
title: "Elevational range shifts of mountain bumble bees and their floral hosts will be punctuated by nonlinear turnover of species and interactions"

@@

Climate-driven range shifts are a fundamental evolutionary process, but the rapid pace of recent climate change has made biogeographic dynamics as much a matter of current events as of historic speculation. This provides a unique opportunity to study biogeographic processes in real time, but it also poses a grave threat to global biodiversity as species are forced to respond rapidly within the constraints of gene pools shaped by cooler climates. Climate-induced range shifts involve more than merely tracking suitable temperature envelopes; when species move, they do so in the context of biological communities and under the influence of more than just temperature, and the outcomes of climate-induced range shifts are thus contingent on potentially nonlinear processes that cannot be reduced to simple elevational or latitudinal temperature gradients. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and disrupted pollination networks. Interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists upon which climate change will act. In this study, we demonstrate that mountain bumble bees and flora are distributed in strongly nonlinear patterns in space and time, and that these patterns propagate from the level of species abundances to that of β-diversity, and from the level of community composition to that of biotic interactions. In particular, we show that the tree line ecotone is a critical threshold of floral resource availability, a distributional interface between high- and low-elevation bumble bees species, and a zone of rapid acceleration of species and interaction β-diversity. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

@@

The rate of recent climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Species range shifts, however, are not merely abstract, Cartesian translations along axes of latitude, elevation, and temperature; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Moreover, some of these factors, unlike temperature, cannot be assumed to vary linearly with latitude or elevation but instead exhibit marked nonlinearities and thresholds. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir]. Thus, species range shifts --- and, perhaps more importantly, the ecological consequences thereof --- are contingent upon existing patterns of species distributions and biotic interactions, and community ecology is as much prior as posterior to the ecology of global change.

Bumble bees (Hymenoptera: *Bombus* spp.) are both classic model taxa in community ecology and an emergent conservation priority due to recent declines [@Goulson2008-oe] and manifest vulnerability to climate change [@Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya), and its subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, bumble bees remain characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity associated with mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective, but they also provide a model system in which to study the general phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, hinge upon an understanding the current elevational and temporal patterns of the bumble bee and floral communities that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we move to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most severe

@@

The study was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management. The sites sampled in this study are the same as those used in a series of previous studies [@Hoiss2012-sn; @Hoiss2012-kr; @Hoiss2015-gv], but the data reported here are otherwise independent.

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected on dry days when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on or a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.1 m^2^. Species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied markedly,  with 153 in 2010, 212 in 2011, and 539 in 2012. The increased total sampling effort in 2012 was driven in part by the overall scarcity of bumble bees that year. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix X for details of sampling patterns.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date (excluding males). To represent floral abundance from the perspective of each bumble bee species in our network, we first scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of visited genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be distinguished reliably in the field. Similarly, we pooled representatives of the parasitic *Psithyrus* subgenus --- which included *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into a single species group *B. psithryus*.

@@

All analyses were conducted in `R` `r citep(citation())`. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa], and the models we fit correspond to the “type I” model form described in @Pedersen2019-kf. PCoA and species β-diversity calculation were performed with the package `vegan` `r citep(citation("vegan"))`. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` `r citep(citation("gdm"))`. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("gdm"))`. Annotated R code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

@@

The remaining species exhibited relatively diffuse patterns of abundance. Of these, *B. soroeensis* was exceptional in that it exhibited peak abundance at the highest sites, but it was also found at relatively high abundance across the whole elevation gradient, peaking at mid-season. *B. hortorum* and *B. lapidarius* peaked at low elevation. *B. jonellus* exhibited little elevational variation in abundance until declining above 1500 m, but its abundance was strongly patterned in time, peaking early in the season. *B. terrestris/lucorum* was generally abundant across the whole elevation gradient and throughout the whole season, but it showed a weakly trimodal pattern in elevation, with abundance peaks at low, middle, and high sites. Its strongest abundance peak occurred at low elevation, early in the season during colony-founding. *B. wurflenii* was broadly abundant above 1000 m but was rarer at lower sites.

Elevation-time GAM smooths of floral abundance differed significantly from a flat surface for all species (p << 0.00001) and explained 64.8% of total deviance (adjusted R^2^ = 0.425). Floral resource availability exhibited similar patterns for most bumble bee species, with peak floral resources occurring below 1500 m and during mid-season **(Figure 2)**. There was a noticeable temporal lag in floral resource availability associated with increasing elevation, consistent with expected temperature constraints on plant phenology. The only species to exhibit marked anomalies in floral resource availability was *B. gerstaeckeri*, which specializes on the genus *Aconitum* [@Ponchau2006-af] but was also observed (presumably collecting nectar) on *Gentiana*, *Salvia*, *Rhinanthus*, *Cirsium*, and *Clinopodium.* Its floral resources were concentrated in two discrete peaks, the first at low elevation and mid-season, and the second at high elevation and late-season.

@@

```{r fig4, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

@@

@@

Arguably the most salient pattern in our data is the unique importance of the elevation band falling at about 1500 m, which corresponds to the lower edge of the tree line ecotone in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn] and specifically that of the elevational transects used in our study. The floristic importance of the tree line ecotone has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id]. By studying the floral community through the lens of foraging bumble bees, however, we provide a functional extension of this floristic pattern, demonstrating that the tree line ecotone comprises (1) a boundary above which floral resource availability sharply declines, (2) a distributional interface between high-elevation bumble bees and their low-elevation or habitat-generalist counterparts, and (3) a zone of rapid acceleration of species and interaction β-diversity. The significance of the tree line ecotone as a threshold of floral resource availability and as a zone of overlapping bumble bee distributions corroborates the recent findings of @Minachilis2020-xd in the Mediterranean system of Mount Olympus, indicating that these patterns are likely to generalize well across mountain ecosystems. Our finding that interaction β-diversity is also punctuated by the tree line ecotone follows naturally from these compositional patterns, and we expect that the β-diversity patterns found in our system will prove similarly generalizable.

The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests that high elevation bumble bees could move upslope faster than their floral resources, though the opposite pattern could also occur if the climate-tracking of bumble bees is limited by other factors, such as nest site availability [@Marshall2020-pc]. A somewhat unexpected finding of our study is that high-elevation bumble bee species (*B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus*) were associated with floral hosts that peaked in abundance at much lower elevations, largely outside the range of their bumble bee visitors. As the upper boundary of high floral resource abundance, the lower edge of the tree line ecotone likely functions as critical foraging habitat for high-elevation bumble bees. Future warming could force these species farther into resource-poor alpine grasslands while at the same time increasing resource competition at the tree line as dominant low-elevation species, like *B. pascuorum*, extend their ranges upslope.

Our analyses of species and interaction β-diversity reveal a striking contrast between a relative stable bumble bee community and an extremely dynamic floral community. The dominance of total interaction β-diversity by floral species turnover is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps a predictable relationship given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that turns over rapidly in both space and time. Indeed, the remarkable flexibility and intelligence of bumble bee foraging [@Heinrich1979-ep; @Loukola2017-bt], which has been interpreted as a side-effect of their sociality [@Dukas1991-ig], could be interpreted alternatively as an adaptation to precisely this problem of foraging in the context of extreme β-diversity. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df].

@@

Mountains are unique ecological theaters in which extrinsic temporal climate change interacts with intrinsic elevational climate gradation to alter the spatial and temporal distributions of species and interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj]. The central finding of our study, that elevation gradients are characterized by nonlinear responses of species, communities, and interaction networks --- punctuated dramatically by the tree line ecotone --- is almost certainly not an idiosyncrasy of bumble bees and their floral hosts but rather a pattern emerging from fundamental physical constraints that should be expected to obtain broadly across the biotic communities of mountains worldwide. Species interactions along tree line ecotones, therefore, deserve special consideration in research and conservation management, since climate-induced range shifts at the tree line ecotone should be expected to have disproportionate ecological consequences relative to range shifts at other elevation bands.

<!-- Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees [@Hines2008-er; @Martinet2018-rb]. Nevertheless, the rapid pace of recent climate change, together with the interacting effects of human land use, are predicted to [@Soroye2020-zb; @Sirois-Delisle2018-vi]  -->

<!-- - What do these findings tell use about climate change and range shifts in mountain ecosystems more generally? -->
<!-- - Do our results suggest any management strategies? -->

<!-- Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees, but the rate of modern climate change, together with the interacting effects of human land use, are predicted to [@Soroye2020-zb; @Sirois-Delisle2018-vi]  -->

<!-- though as a matter of applied conservation it is perhaps doubtful that local interventions could forestall for long the consequences of a process occurring at a global scale. -->
